A Generalized Sphere Theorem for Positively Curved Combinatorial 3-Manifolds

### 2013-06-14 paper/paper.tex
more writing. first draft done

Why are these paths relevant to this paper? In the classical setting, the generalized sphere theorem requires a manifold to have diameter more than half the maximum allowed by the Bonnet-Myers theorem, and this diameter bound cannot be improved. If we naively imitate the classical results in the discrete setting, from Theorem \ref{thm:discrete_BM} we would want to assume the bound $diam_1(M)>\frac{5}{2}$. Right away we see a discrepancy between the discrete and classical results since this bound cannot possibly be sharp. Worse, the smallest weakening that could be sharp is $diam_1(M) \geq 3$ and this bound does not work. As we will see, there are positively curved combinatorial 3-manifolds $M$ with $diam_1(M)=3$ but which are {\em not} homeomorphic to the 3-sphere. However, if we instead use the finer-grained measure of diameter, $diam(M)$, the problem disappears and the discrete results mirror the classical smooth ones exactly. That is, we can prove a discrete version of the generalized sphere theorem with a diameter bound exactly half the maximum allowed by the corresponding discrete Bonnet-Myers theorem.

@@

To see this, consider a combinatorial 3-manifold $M$ containing a minimal jump from vertex $v$ to vertex $w$ passing through edges $e_1$ and $e_2$ (as seen in Figure \ref{fig:jump}.) Using the structure of a jump, we will construct two edge-paths starting at $v$. We choose the first edge from $v$ to some vertex in $e_1$ and then we can continue the path to either of the vertices in $e_2$. The resulting paths, $Q_1$ and $Q_2$, must be minimal among edge-paths. If they were not, then their endpoints would be connected by an edge and a two-edge path from $v$ to $w$ would exist, contradicting the minimality of the jump. Thus $Q_1$ and $Q_2$ provide examples of minimal edge-paths which ``branch''.

The non-branching property of minimal paths is restored, at least for positively curved manifolds, if we include hops and jumps. See \cite{Trout10}. In particular we have:

@@

\noindent It is not surprising that the

>>>>

Why are these paths relevant to this paper? In the classical setting, the generalized sphere theorem requires a manifold to have diameter more than half the maximum allowed by the Bonnet-Myers theorem, and this diameter bound cannot be improved. If we naively imitate the classical results in the discrete setting, from Theorem \ref{thm:discrete_BM} we would want to assume the bound $diam_1(M)>\frac{5}{2}$. Right away we see a discrepancy between the discrete and classical results since this bound cannot possibly be sharp. Worse, the smallest weakening that could be sharp is $diam_1(M) \geq 3$ and this bound does not work. As we will see, there are positively curved combinatorial 3-manifolds $M$ with $diam_1(M)=3$ but which are {\em not} homeomorphic to the 3-sphere. However, if we instead use the finer-grained measure of diameter, $diam(M)$, the problem disappears and the discrete results mirror the classical ones exactly. That is, we can prove a discrete version of the generalized sphere theorem with a diameter bound exactly half the maximum allowed by the corresponding discrete Bonnet-Myers theorem.

@@

To see this, consider a combinatorial 3-manifold $M$ containing a minimal jump from vertex $v$ to vertex $w$ passing through edges $e_1$ and $e_2$ (as seen in Figure \ref{fig:jump}.) Using the structure of a jump, we will construct two edge-paths starting at $v$. We choose the first edge from $v$ to some vertex in $e_1$ and then we can continue the path to either of the vertices in $e_2$. The resulting paths, $Q_1$ and $Q_2$, must be minimal among edge-paths. If they were not, then their endpoints would be connected by an edge and a two-edge path from $v$ to $w$ would exist, contradicting the minimality of the jump. Thus $Q_1$ and $Q_2$ provide examples of minimal edge-paths which ``branch''. The non-branching property of minimal paths is restored, at least for positively curved manifolds, if we include hops and jumps. In particular we have:

@@

\noindent See \cite{Trout10} for details.

It is not entirely surprising that discrete differential geometry on a PL-manifold $M$ comes to resemble classical differential geometry more and more as the set of admissible PL-paths on $M$ is expanded. Forcing ourselves to use only edges amounts to coarse-graining the underlying PL-geometry, and expanding the set of paths clearly reduces this coarse-graining. We could even include {\em all} the PL-paths in $M$ and still have a ``discrete'' geometry in some sense, because our PL-metric is determined solely by the structure of $M$ as a simplicial complex. However, since PL-paths admit infinitesimal deformations we would lose much of the ``discrete'' flavor to the theory. The successful use of hops and jumps in this paper and in \cite{Trout10} indicate that one can improve the behavior of discrete geometries by judicious inclusion of PL-paths in addition to edges. Investigating the possibilities for such inclusions and their benefits represents an exciting topic of further research.
